A prediction and Analysis of the interplay between Graduate Statistics and Salary Prospects in the STEM Field

### 2024-01-29 doc/DataLiteracyStudentProject/paper.tex
updates

\twocolumn[
\icmltitle{All just Greedy for Money? - A prediction and Analysis of the interplay between Graduate Statistics and Salary Prospects in the STEM Field}

@@

% fill in your matrikelnummer, email address, degree, for each group member
\icmlaffiliation{first}{Matrikelnummer 5977981, abdallah.abdul-latif@student.uni-tuebingen.de, MSc Computer Science}
\icmlaffiliation{second}{Matrikelnummer 4189024, lisa-maria.fritsch@student.uni-tuebingen.de, MSc Computer Science}
\icmlaffiliation{third}{Matrikelnummer 12345678, first.last@student.uni-tuebingen.de, MSc Media Informatics}
\icmlaffiliation{fourth}{Matrikelnummer 6040570, maximilian.schnitt@student.uni-tuebingen.de, BSc Computer Science}

@@

@@

\subsection{Model creation}\label{sec:methods:var}
A vector autoregressive model is used to analyze relationships between multiple time series.
The model creates a linear combination for each prediction which can be expressed in a matrix-vector notation.
\begin{equation*}
    \resizebox{\linewidth}{!}{
        $\begin{bmatrix}
            S_{1,n}\\
            S_{2,n}\\
            \vdots \\
            S_{m,n}
        \end{bmatrix} = 
        \begin{bmatrix}
            C_{1,1} & C_{1,2} & \dots &C_{1,n-1}\\
            C_{2,1} & C_{2,2} &\dots & C_{2,n-1}\\
            \vdots & \vdots  & \ddots &\vdots\\
            C_{m,1} & C_{m,2} &\dots &C_{m,n-1}\\
        \end{bmatrix} \cdot
        \begin{bmatrix}
            S_{1,1} & S_{1,2} & \dots & S_{1,n-1}\\
            S_{2,1} & S_{2,2} & \dots & S_{2,n-1}\\
            \vdots & \vdots  & \ddots &\vdots\\
            S_{m,1} & S_{m,2} & \dots & S_{m,n-1}\\
        \end{bmatrix}+
        \begin{bmatrix}
            \epsilon_{1}\\
            \epsilon_{2}\\
            \vdots\\
            \epsilon_{n}\\
        \end{bmatrix}$%
    }
\end{equation*}
The $\epsilon$ vector approximates an error, which is induced by the model. One main benefit of the VAR model is that it also assumes
that $S_{i,j}$ is also influenced by $S_{k,l}$ $i,k \in \{1,\dots, m\}: i\neq k$ and $j,l \in \{1,\dots, n-1\}: j\neq l$.
A VAR model has one major advantage, over other possible approaches, which is its simplicity. Therefore its results remain interpretable.
This model only uses one hyperparameter which describes the amount of past data that is used to make a prediction. This parameter is called the lag.
The vector autoregressive model heavily depends on the stationarity assumption which ensures that the stochastical process stays the same over time. 
In this context, this is assured by the following three axioms:
\begin{enumerate}[label=\Roman*]
    \item The mean $\mu$ is constant
    \item The standard deviation is constant
    \item There is no seasonality in the data
\end{enumerate}
Since we are only interested in predicting two time steps into the future, which is equivalent to one year,
we can safely assume that our data is stationary.

\subsection{Choosing lag}\label{sec:methods:lag}
To find the optimal hyperparameter we used two well-established methods, the PACF and information-based criteria. Particularly the AIC and 
BIC. The PACF only considers the impact of the previous data point $S_{n-1}$ on $S_{n}$ to build a $(1-\alpha)$ confidence interval.
We choose $\alpha = 5\%$ and get get a 95\% Confidence Interval 
\begin{table}[H]
    \centering
    \begin{tabular}{c|c|c|c}
        & students & salary & graduates \\
        \hline
        Prediciton 1 & 1.257\% & 1.271\% & 3.944\% \\
        \hline
        Prediciton 2 & 4.843\% & 0.119\% & 1.722\%
    \end{tabular}
    \caption{Difference in percent between the true value and predicted value for students, salary, and graduates. Lag set to $1$.}
\end{table}
The percentage differences between the true values especially for the students are quite high. To get better results
the AIC and BIC are used to find a better lag for the hyperparameter $p$ \textcolor{red}{is $p$ = lag?}. The abbreviation AIC
stands for Akaike Information Criterion and tries to find an optimum between the complexity of the model and fitting the data.
This should prevent overfitting. 
BIC is short for Bayesian Information Criterion and penalizes more complex models.
Therefore BIC leads to less complicated models than the ones from AIC, which sets the two apart.
The optimal lag based on AIC is 6, while for BIC it is 1, which is consistent with the PACF.
\begin{table}[H]
    \centering
    \begin{tabular}{c|c|c|c}
        & students & salary & graduates \\
        \hline
        Prediciton 1 & 0.484\% & 0.419\% & 0.111\% \\
        \hline
        Prediciton 2 & 1.346\% & 0.306\% & 0.939\%
    \end{tabular}
    \caption{Difference in percent between the true value and predicted value for students, salary, and graduates. Lag set to $5$.}
\end{table}

The percentage errors show that there is a significant improvement in the prediction of student numbers compared to the lag order of 1.
Setting lag to $5$ results in even better errors for the domains of our interest, predicting students.

\subsection{Test for causality}\label{sec:methods:causality}
To ensure that the model does not try to find structure in the data, where there is none, we applied the Granger causality test.
Granger causal implies that one time series is the result of another shifted time series.
In this case, we try to find out if e.g. the salary time series is just a shifted version of the student's or the graduate's time series.
The test does verifies this by using series $S_i$ and trying to insert data points from $S_j$ or $S_k$ $i \neq j\neq k$
to predict the outcome of the model. After inserting the data points a t-test and a F-test are performed. If both
tests return statistical significance, then we showed that $S_i$ could be expressed by a shifted input time series $S_j$ with $j\neq i$.
\textcolor{red}{OK, and did we show that?}

\subsection{Fitted model}\label{sec:methods:fit}
\begin{figure}
    \includegraphics{fig/params_Students.pdf}
    \caption{Coefficients of VAR model to predict students time-series
    for $t_0$. $t_{-1}$ to $t_{-5}$ are the values of each time series respectively at the lagged time step.
    $t_{-1}$ is the most recent value.}
    \label{fig:student coefficients}
\end{figure}

@@

To further understand the data underlying correlations, one can look at the correlation matrix
of residuals in \Cref{table:correlation matrix}.
This matrix describes the degree to which the model's errors, called residuals, are related to each other.
High correlations between errors can indicate a not fully captured underlying structure.
Values close to zero, like $-0.115$ between salary and graduates, indicate that the errors
in predicting the salary variable are not related to the errors in predicting the graduates variable.

\begin{table}[H]
    \centering
    \begin{tabular}{c|c|c|c}
        & students & salary & graduates\\
        \hline
        students & 1.000 & -0.115 &  0.111\\
        \hline
        salary & -0.115 & 1.000000 & -0.902\\
        \hline
        graduates & 0.111 & -0.902 &  1.000\\
    \end{tabular}
    \caption{Correlation matrix of residuals using $5$ lags.}
    \label{table:correlation matrix}
\end{table}

@@

\section*{Contribution Statement}
\textcolor{red}{TODO}
Abdallah Abdul-Latif analyzed the optimal parameter lag and mapped courses to salary sectors.
Lisa-Maria Fritsch
Paul Kaifler 
Maximilian Schnitt parsed university data and analyzed optimal parameter lag.
All authors jointly did the analyses of the data and wrote the text of the report.
% Max Mustermann collected and prepared data. Gabi Musterfrau and John Doe performed the data analysis. Jane Doe produced visualizations. All authors will jointly wrote the text of the report. Note that you, as a group, a collectively responsible for the report. Your contributions should be roughly equal in amount and difficulty.

@@

>>>>

\twocolumn[
\icmltitle{A prediction and Analysis of the interplay between Graduate Statistics and Salary Prospects in the STEM Field}

@@

% fill in your matrikelnummer, email address, degree, for each group member
\icmlaffiliation{first}{Matrikelnummer 5977981, abdallah.abdul-latif@student.uni-tuebingen.de, MSc Computer Science}
\icmlaffiliation{second}{Matrikelnummer 4189024, lisa-maria.fritsch@student.uni-tuebingen.de, MSc Computer Science}
\icmlaffiliation{third}{Matrikelnummer 5993286, paul.kaifler@student.uni-tuebingen.de, Bsc Computer Science}
\icmlaffiliation{fourth}{Matrikelnummer 6040570, maximilian.schnitt@student.uni-tuebingen.de, BSc Computer Science}

@@

A repository containing the complete analyses and results can be found \href{https://github.com/mxs01/AnalysingStudentDevelopment}{here}.

@@

\subsection{Model creation}\label{sec:methods:var}
A vector autoregressive model is used to analyze relationships between multiple time series.
The model creates a linear combination for each prediction which can be expressed in a matrix-vector notation.
\begin{equation*}
    \resizebox{\linewidth}{!}{
        $\begin{bmatrix}
            S_{1,n}\\
            S_{2,n}\\
            \vdots \\
            S_{m,n}
        \end{bmatrix} = 
        \begin{bmatrix}
            C_{1,1} & C_{1,2} & \dots &C_{1,n-1}\\
            C_{2,1} & C_{2,2} &\dots & C_{2,n-1}\\
            \vdots & \vdots  & \ddots &\vdots\\
            C_{m,1} & C_{m,2} &\dots &C_{m,n-1}\\
        \end{bmatrix} \cdot
        \begin{bmatrix}
            S_{1,1} & S_{1,2} & \dots & S_{1,n-1}\\
            S_{2,1} & S_{2,2} & \dots & S_{2,n-1}\\
            \vdots & \vdots  & \ddots &\vdots\\
            S_{m,1} & S_{m,2} & \dots & S_{m,n-1}\\
        \end{bmatrix}+
        \begin{bmatrix}
            \epsilon_{1}\\
            \epsilon_{2}\\
            \vdots\\
            \epsilon_{n}\\
        \end{bmatrix}$%
    }
\end{equation*}
The $\epsilon$ vector approximates an error, which is induced by the model. One main benefit of the VAR model is that it also assumes
that $S_{i,j}$ is also influenced by $S_{k,l}$ $i,k \in \{1,\dots, m\}: i\neq k$ and $j,l \in \{1,\dots, n-1\}: j\neq l$. As mentioned in \citep{kot}
A VAR model has one major advantage, over other possible approaches, which is its simplicity. Therefore its results remain interpretable.
This model only uses one hyperparameter which describes the amount of past data that is used to make a prediction. This parameter is called the lag.
The vector autoregressive model heavily depends on the stationarity assumption which ensures that the stochastical process stays the same over time. 
In this context, this is assured by the following three axioms:
\begin{enumerate}[label=\Roman*]
    \item The mean $\mu$ is constant
    \item The standard deviation is constant
    \item There is no seasonality in the data
\end{enumerate}
Since we are only interested in predicting two time steps into the future, which is equivalent to one year,
we can safely assume that our data is stationary.

\subsection{Choosing lag}\label{sec:methods:lag}
To find the optimal hyperparameter we used two well-established methods, the PACF and information-based criteria. Particularly the AIC and 
BIC. The PACF only considers the impact of the previous data point $S_{n-1}$ on $S_{n}$ to build a $(1-\alpha)$ confidence interval.
We choose $\alpha = 5\%$ and get a 95\% Confidence Interval. The optimal value according to PACF is $3$.
\begin{table}[H]
    \centering
    \begin{tabular}{c|c|c|c}
        & students & salary & graduates \\
        \hline
        Prediciton 1 & 0.704\% & 0.871\% & 2.658\% \\
        \hline
        Prediciton 2 & 2.786\% & 0.301\% & 0.286\%
    \end{tabular}
    \caption{Difference in percent between the true value and predicted value for students, salary, and graduates. Lag set to $3$.}
    \label{tab:resid1}
\end{table}
The percentage differences between the true values and the predicted ones, especially for the students, are quite high, see \Cref{tab:resid1}.
To get better results
the AIC and BIC are used to find a better lag for the hyperparameter $p$, which is the lag. The abbreviation AIC
stands for Akaike Information Criterion and tries to find an optimum between the complexity of the model and fitting the data.
This should prevent overfitting. 
BIC is short for Bayesian Information Criterion and penalizes more complex models.
Therefore BIC leads to less complicated models than the ones from AIC, which sets the two apart.
Both BIC and AIC return $6$ as an optimal lag value for our data set. Since we are focusing on predicting students, low errors in that domain
are more important to us. Minimizing these errors within the range from $3$ to $6$ leads to an optimal lag value of $5$.
\begin{table}[H]
    \centering
    \begin{tabular}{c|c|c|c}
        & students & salary & graduates \\
        \hline
        Prediciton 1 & 0.484\% & 0.419\% & 0.111\% \\
        \hline
        Prediciton 2 & 1.346\% & 0.306\% & 0.939\%
    \end{tabular}
    \caption{Difference in percent between the true value and predicted value for students, salary, and graduates. Lag set to $5$.}
    \label{tab:resid2}
\end{table}

The percentage errors in \Cref{tab:resid2} show that there is a significant improvement in the prediction of student numbers
compared to the lag order of $3$ from \Cref{tab:resid1}.

\subsection{Test for causality}\label{sec:methods:causality}
To ensure that the model does not try to find structure in the data, where there is none, we applied the Granger causality test.
Granger causal implies that one time series is the result of another shifted time series.
In this case, we try to find out if e.g. the salary time series is just a shifted version of the student's or the graduate's time series.
The test does verifies this by using series $S_i$ and trying to insert data points from $S_j$ or $S_k$ $i \neq j\neq k$
to predict the outcome of the model. After inserting the data points a t-test and a F-test are performed. If both
tests return statistical significance, then we showed that $S_i$ could be expressed by a shifted input time series $S_j$ with $j\neq i$.
% tbd notebook 2
\textcolor{red}{TBD}

\subsection{Fitted model}\label{sec:methods:fit}
\begin{figure}
    \includegraphics{fig/params_Students.pdf}
    \caption{Coefficients of VAR model to predict students time-series
    for $t_0$. $t_{-1}$ to $t_{-5}$ are the values of each time series respectively at the lagged time step, which is equal to one semester.
    $t_{-1}$ is the most recent value.}
    \label{fig:student coefficients}
\end{figure}

@@

To further understand the data underlying correlations, one can look at the correlation matrix
of residuals in \Cref{tab:correlation matrix}.
This matrix describes the degree to which the model's errors, called residuals, are related to each other.
High correlations between errors can indicate a not fully captured underlying structure.
Values close to zero, like $-0.115$ between salary and graduates, indicate that the errors
in predicting the salary variable are not related to the errors in predicting the graduates variable.

\begin{table}[H]
    \centering
    \begin{tabular}{c|c|c|c}
        & students & salary & graduates\\
        \hline
        students & 1.000 & -0.115 &  0.111\\
        \hline
        salary & -0.115 & 1.000000 & -0.902\\
        \hline
        graduates & 0.111 & -0.902 &  1.000\\
    \end{tabular}
    \caption{Correlation matrix of residuals using $5$ lags.}
    \label{tab:correlation matrix}
\end{table}

@@

\section*{Contribution Statement}
Abdallah Abdul-Latif analyzed the optimal parameter lag and mapped courses to salary sectors.
Lisa-Maria Fritsch was responsible for data pre-processing.
Paul Kaifler fitted the model to the data and created visualizations.
Maximilian Schnitt parsed the university data and analyzed optimal parameter lag.
All authors jointly did the analyses of the data and wrote the text of the report.
% Max Mustermann collected and prepared data. Gabi Musterfrau and John Doe performed the data analysis. Jane Doe produced visualizations. All authors will jointly wrote the text of the report. Note that you, as a group, a collectively responsible for the report. Your contributions should be roughly equal in amount and difficulty.

@@

% References
